Tropical mathematics and the Lambda-calculus I

### 2023-08-02 CSL24/main.tex
commit

%TODO mandatory: add short abstract of the document
\begin{abstract}
We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, and we show that it provides a unified framework for both program  metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation. 
To do that we focus on the semantics arising from the relational model weighted over the tropical semiring, and we discuss its application to the study of “best case” program behavior for languages with probabilistic and non-deterministic effects. Finally, we show that a general foundation for this approach is provided by an abstract correspondence between tropical algebra and Lawvere’s theory of generalized metric spaces.
%A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
%We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
\end{abstract}

>>>>

%TODO mandatory: add short abstract of the document
\begin{abstract}
We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, and we show that it provides a unifying framework for two well-developed quantitative approaches to program semantics: on the one hand program metrics, based on the analysis of program sensitivity via Lipschitz conditions, on the other hand resource analysis, based on linear logic and higher-order program differentiation. 
To do that we focus on the semantics arising from the relational model weighted over the tropical semiring, and we discuss its application to the study of “best case” program behavior for languages with probabilistic and non-deterministic effects. Finally, we show that a general foundation for this approach is provided by an abstract correspondence between tropical algebra and Lawvere’s theory of generalized metric spaces.
%A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
%We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
\end{abstract}

### 2023-08-02 CSL24/section1.tex
commit

In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like computing the number of computation steps or convergence probabilities, 
as opposed to purely \emph{qualitative} properties like termination or program equivalence. 
Notably, a significant effort has been made to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties. We can mention, for example, 
intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022} or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for 
probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex programs from the properties of their constituent parts.

@@

On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}. This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.

@@

A natural question, at this point, is whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled. At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a {linear} (hence \emph{Lipschitz}) function $n x$, while in differential models this results in a {polynomial} function $x^{n}$, thus not Lipschitz. The fundamental intuition of this work is then the observation that 
this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n x$.

@@

Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as an alternative approach to algebra and geometry where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
%
%
% interpreting the usual ``$\times$'' and  ``$+$'' operations by  ``$+$'' by ``$\min$''. It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
%Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
%played by the idempotent operation $\min$.
For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
$
\varphi(x,y)=\min\{2x, x+2y, 3y\}
$.
%This is not a \emph{ad-hoc} setting: 
In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Sturmfelds}.
Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (and, moreover,  the tropical roots can be used to approximate the actual roots \cite{Noferini2015}).
The computational nature of tropical notions explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).

Coming back to our discussion on program semantics, tropical mathematics seems to be just  what we look for, as it turns polynomial functions like $x^{n}$ into Lipschitz maps like $n x$.
At this point, it is worth mentioning that a tropical variant of the usual relational semantics of linear logic and the $\lambda$-calculus has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.

In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs.% (an aspect that will be developed in more detail in a future paper of this series).

@@

\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ (simply typed $\lambda$-calculus) and $\mathrm{PCF}$. Notably, 
 we show that higher-order programs are interpreted by a  
generalizations of {tropical power series} \cite{Porzio2021}, and we show that these functions are locally Lipschitz-continuous, thus yielding a full-scale metric semantics. This is in Sections \ref{section3} and~\ref{sec:tls}.

@@

\item We conclude 
by putting the connection between the 
tropical, differential and metric viewpoints at the right level of abstraction and generality.
We recall a well-known correspondence between Lawvere's \emph{generalized metric spaces} \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007} and we show that it yields a model of the differential $\lambda$-calculus which extends the tropical relational model. This is in Section~\ref{sec:GMS}.
\end{itemize}
%
%\section{Bounded and Differential $\lambda$-Calculi}
%
%
%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
%$$
%A::= o \mid !_{n}A \multimap A
%$$
%
%
%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
%$$
%A::= o \mid [A, \dots , A] \multimap A
%$$
%
%
%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
%
%\begin{proposition}
%$\Gamma \vdash_{\BSTLC} M:A$ implies 
%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
%\end{proposition}
%

>>>>

In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like e.g.~the number of computation steps or the probability of convergence, 
as opposed to purely \emph{qualitative} properties like termination or program equivalence. 
Notably, a significant effort has been made to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties. We can mention, for example, 
intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022} or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for 
probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex programs from the properties of their constituent parts.

@@

On the one hand one one could mention the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}. This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.

@@

A natural question, at this point, is whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled. At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models a program duplicating an input $n$ times yields a {linear} (hence \emph{Lipschitz}) function $n x$, whereas in differential models it would lead to a {polynomial} function $x^{n}$, thus not Lipschitz. The fundamental idea behind this work is the observation that 
this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we will see, the monomial $x^{n}$ precisely reads as the linear function $n x$.

@@

Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as an alternative approach to algebra and geometry where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
%
%
% interpreting the usual ``$\times$'' and  ``$+$'' operations by  ``$+$'' by ``$\min$''. It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
%Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
%played by the idempotent operation $\min$.
For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
$
\varphi(x,y)=\min\{2x, x+2y, 3y\}
$.
%This is not a \emph{ad-hoc} setting: 
In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Sturmfelds}.
Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (and, moreover,  the tropical roots can be used to approximate the actual roots \cite{Noferini2015}).
The combinatorial nature of several methods in tropical mathematics explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).

Coming back to our discussion on program semantics, tropical mathematics seems to be just  what we look for, as it turns polynomial functions like $x^{n}$ into Lipschitz maps like $n x$.
At this point, it is worth mentioning that a tropical variant of the usual relational semantics of linear logic and the $\lambda$-calculus has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics and of the potentialities of its applications has not yet been undertaken.

In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and suggests new combinatorial methods to study probabilistic and non-deterministic programs.% (an aspect that will be developed in more detail in a future paper of this series).

@@

\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of the simply typed $\lambda$-calculus ($\STLC$ in the following) and $\mathrm{PCF}$ \cite{Plotkin1977}. Notably, 
 we show that higher-order programs are interpreted by a  
generalizations of {tropical power series} \cite{Porzio2021}, and we show that these functions are locally Lipschitz-continuous, thus yielding a full-scale metric semantics. This is in Sections \ref{section3} and~\ref{sec:tls}.

@@

\item We conclude 
by framing the connection between the 
tropical, differential and metric viewpoints at a more abstract level.
We recall a well-known correspondence between Lawvere's \emph{generalized metric spaces} \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007} and we show that it yields a model of the differential $\lambda$-calculus which extends the tropical relational model. This is in Section~\ref{sec:GMS}.
\end{itemize}
%
%\section{Bounded and Differential $\lambda$-Calculi}
%
%
%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
%$$
%A::= o \mid !_{n}A \multimap A
%$$
%
%
%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
%$$
%A::= o \mid [A, \dots , A] \multimap A
%$$
%
%
%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
%
%\begin{proposition}
%$\Gamma \vdash_{\BSTLC} M:A$ implies 
%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
%\end{proposition}
%

### 2023-08-02 CSL24/section4.tex
commit

%Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
The following proposition provides a useful characterization of the functional metrics in $\LREL$, relying on 
the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
Remember that $d_\infty(f,g):=\sup_{a\in X} d_\infty(f(x),f(y))$.

\begin{proposition}
For all $f,g\in \LREL_{!}(X,Y)$, $d_{\infty}( f,g)=d_\infty(f^{!},g^{!})$.% $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}} \norm{ f( x)-g(x)}_{\infty}$.
\end{proposition}

@@

Observe that, for any tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \overline{\BB R}_{\geq 0}$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$). Hence, using Proposition \ref{prop:troplinear} 2., we have:
\begin{corollary}\label{prop:polylip}
Any tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to\overline{\BB R}_{\geq 0}$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
\end{corollary}

>>>>

%Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
Observe that on the hom-sets $\HOM{\LREL_{!}}{X}{Y}$ there are two natural notions of distance: the metric $\| f-g\|_{\infty}$ arising from the norm and the one arising from the usual $\sup$-metric $d_\infty(f,g):=\sup_{a\in X} d_\infty(f^{!}(x),f^{!}(y))$.
In general one has $\| f-g\|_{\infty}\geq d_\infty(f,g)$, the equality holding when $f,g$ are linear.

%The following proposition provides a useful characterization of the functional metrics in $\LREL$, relying on 
%the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
%Remember that $d_\infty(f,g):=\sup_{a\in X} d_\infty(f(x),f(y))$.
%
%\begin{proposition}
%For all $f,g\in \LREL_{!}(X,Y)$, $d_{\infty}( f,g)=d_\infty(f^{!},g^{!})$.% $\norm{ \matr f-\matr g}_{\infty} =  \sup_{x\in \Lawv^{X}} \norm{ f( x)-g(x)}_{\infty}$.
%\end{proposition}

@@

For any tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to \overline{\BB R}_{\geq 0}^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \overline{\BB R}_{\geq 0}$ (as a monomial $\mu_ix+c_{i}$ yields a matrix entry on $!_{\#\mu_i}X\times Y$). Hence, using Proposition \ref{prop:troplinear} 2., we have:
\begin{corollary}\label{prop:polylip}
Any tropical polynomial $\varphi:\overline{\BB R}_{\geq 0}^{X}\to\overline{\BB R}_{\geq 0}$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
\end{corollary}

### 2023-08-02 CSL24/section5bis.tex
commit

\subparagraph*{Metric approach: bounded $\lambda$-calculus% as Lipschitz functions
}

@@

\subparagraph*{Resource approach: the differential $\lambda$-calculus% as polynomials
}

@@

\subparagraph*{Tropical Mathematics: a possible synthesis}%Outline of the paper}

>>>>

\subparagraph*{Metric Approach: Bounded $\lambda$-Calculus% as Lipschitz functions
}

@@

\subparagraph*{Resource Approach: the Differential $\lambda$-Calculus% as polynomials
}

@@

\subparagraph*{Tropical Mathematics: Lipschitz Meets Taylor}%Outline of the paper}

### 2023-08-02 CSL24/sectionTaylip.tex
commit

\begin{corollary}\label{cor:taylor}
Suppose $x: A \vdash_{\STLC}M:B$ and $\vdash_{\STLC} N:A$. 
Then for all $t\in \Te{M}$ such that $\model{t}^{!}(\model{N})\neq \infty$, and $\delta>0$, the tps $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}^{!}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
 $B_{\delta}(\model{N})$.
\end{corollary}
\begin{proof}[Proof sketch.]
We apply the Lipschitz estimate from Thm.~\ref{thmTLSlocLip} for $\model{M}^{!}$,  and use the fact that, from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$.
To conclude, since $\model{t}^{!}$ is concave and non-decreasing, $\max_{\overline{B_{3\delta}(x)}}\model{t}^{!}= \model{t}^{!}(x+3\delta)$, which yields the stated Lipschitz estimate. 
\end{proof}
%{\color{red}ADD PROOF SKETCH}

>>>>

\begin{corollary}\label{cor:taylor}
Suppose $x: A \vdash_{\STLC}M:B$ and $\vdash_{\STLC} N:A$. 
Then for all $t\in \Te{M}$ such that $\model{t}^{!}(\model{N})\neq \infty$, and $\delta>0$, the tps $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}^{!}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
 $B_{\delta}(\model{N})$.
\end{corollary}
\begin{proof}
Thm.~\ref{thmTLSlocLip} yields the estimate $\max_{\overline{B_{3\delta}(\model{N})}}\model{M}^{!}$. As from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$, we deduce that $K=\max_{\overline{B_{3\delta}(\model{N})}}\model{t}^{!}\geq \max_{\overline{B_{3\delta}(\model{N})}}\model{M}^{!}$ is also a local Lipschitz constant for $\model{M}^{!}$. Moreover, since $\model{t}^{!}$ is concave and non-decreasing, the $\max$ of $\model{t}^{!}$ is attained at the maximum point of $\overline{B_{3\delta}(\model{N})}$, that is, 
$K= \model{t}^{!}(x+3\delta)$. Finally, from $\model{t}^{!}(\model{N})<\infty$ and the continuity of $\model{t}^{!}$ we deduce $K<\infty$.
\end{proof}
%\begin{proof}[Proof sketch.]
%We apply the Lipschitz estimate from Thm.~\ref{thmTLSlocLip} for $\model{M}^{!}$,  and use the fact that, from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$.
%To conclude, since $\model{t}^{!}$ is concave and non-decreasing, $\max_{\overline{B_{3\delta}(x)}}\model{t}^{!}= \model{t}^{!}(x+3\delta)$, which yields the stated Lipschitz estimate. 
%\end{proof}
%{\color{red}ADD PROOF SKETCH}
